---
title: "Advanced R Final Project"
author: "Sibi Simon - 18201381"
date: "1 September 2019"
output: html_document
runtime: shiny
---

```{r echo=FALSE, message=FALSE, warning=FALSE}

# Including libraries
library(readr)
library(magrittr)
library(dplyr)
library(ggplot2)
library(ggiraph)
library(viridis)
library(gganimate)
library(rstan)
library(shiny)

# Setting seed to get an unchanged output
set.seed(18201381)
```



#### Dataset

The dataset consist of details of all discovered extra-solar planets. The details of columns are described below.

* 1: Primary identifier of planet
* 2: Binary flag [0=no known stellar binary companion; 1=P-type binary (circumbinary); 2=S-type binary; 3=orphan planet (no star)
* 3: Planetary mass [Jupiter masses]
* 4: Radius [Jupiter radii]
* 5: Period [days]
* 6: Semi-major axis [Astronomical Units]
* 7: Eccentricity
* 8: Periastron [degree]
* 9: Longitude [degree]
* 10: Ascending node [degree]
* 11: Inclination [degree]
* 12: Surface or equilibrium temperature [K]
* 13: Age [Gyr]
* 14: Discovery method
* 15: Discovery year [yyyy]
* 16: Last updated [yy/mm/dd]
* 17: Right ascension [hh mm ss]
* 18: Declination [+/-dd mm ss]
* 19: Distance from Sun [parsec]
* 20: Host star mass [Solar masses]
* 21: Host star radius [Solar radii]
* 22: Host star metallicity [log relative to solar]
* 23: Host star temperature [K]
* 24: Host star age [Gyr]
* 25: A list of lists the planet is on



#### Question 1


```{r message=FALSE, warning=FALSE}
# Reading data using read_csv function
exoplanet_data = read_csv("exo_data.csv")

# Converting columns to approproate type as mentioned
exoplanet_data %<>% mutate(year = as.integer(year), meth = as.factor(meth), flag = as.factor(flag))
```



#### Question 2


```{r}
# Avoiding the exoplanet with unknown method of discovery
exoplanet_data %<>% filter(!is.na(meth))
```



#### Question 3


```{r, warning=FALSE, message=FALSE, fig.align='center'}
# Using ggplot to plot histogram of log distance highlighted by nethod of discovery
ggplot(exoplanet_data, aes(x=log(dist), fill=meth)) +
  geom_histogram(alpha = 0.5, position="dodge") +
  labs(title="Histogram of distance by method of discovery", x = "Distance") +   # Addding labels
  theme_bw()                                                                     # Adding theme
```



#### Question 4

```{r, warning=FALSE, message=FALSE, fig.align='center'}

# Adding the additional column to dataset containig the url to route when clicked
# Url will be base url + id 
exoplanet_data$onclick =  sprintf("window.open(\"%s%s\")",
                        "http://www.openexoplanetcatalogue.com/planet/",
                         as.character(exoplanet_data$id))

# Ploting log distance vs log mass scatterplot and adding interactive feature to it
# The plot is seperated based on the method of discovery
gg_planet = ggplot(exoplanet_data, aes(x = log(dist),
                    y = log(mass), color = meth)) + 
                    geom_point_interactive(aes(data_id = id,           
                    tooltip = id, onclick = onclick)) +                # Adding interactive point
                    facet_grid(. ~ meth, scales = 'free_x') +          # Seperating by method
                    labs(title="Distance vs Mass scatter plot",        # Adding labels
                         x = "Distance", y="Mass") +
                    theme_bw() +                                       # Selecting theme
                    theme(legend.position = 'None')                    # Removing legend

# Adding hover and interactive feature
ggiraph(code = print(gg_planet))
```



#### Question 5


```{r, warning=FALSE, message=FALSE}
# Rename the column name radius into jupiter_radius
exoplanet_data <- rename(exoplanet_data, jupiter_radius = 'radius')

# Computing earth radius
exoplanet_data$earth_radius <- exoplanet_data$jupiter_radius * 11.2
```



#### Question 6


```{r, warning=FALSE, message=FALSE, fig.align='center'}
# Filtering data without NA values in period or earth radius column
exo_planet_data_cluster <- exoplanet_data[complete.cases(exoplanet_data[, c("period", "earth_radius")]),]

# Data with NA in period or earth_radius
exo_planet_data_with_na <- exoplanet_data[!complete.cases(exoplanet_data[, c("period", "earth_radius")]),]

# Clustering using kmeans with 4 centers and nstart 15
km_cl <- kmeans(log(exo_planet_data_cluster %>% select(period, earth_radius)), 
                4, nstart = 15)
```

Here N start 15 is given to converge to global minima rather than the local minima which will give better result than just running once.



#### Question 7


```{r, warning=FALSE, message=FALSE, fig.align='center'}

# Adding cluster as a column in dataset 
exo_planet_data_cluster$type <- km_cl$cluster
# Converting to factor
exo_planet_data_cluster$type <- as.factor(exo_planet_data_cluster$type)

# Plotting the clustered data
ggplot(data=exo_planet_data_cluster, 
       aes(x=log(period), y=log(earth_radius), color=type)) + 
       geom_point() + labs(title="Clustering on Peroid and Earth radius",
                            x = "Period", y="Earth radius") +            # Adding labels  
       theme_bw()                                                        # Adding theme

# Adding an additional factor unknown to type column NA dataset created above
exo_planet_data_with_na$type <- 'unknown'
```


**Comparing with the given picture we can encode the names as given below.**


```{r, warning=FALSE, message=FALSE, fig.align='center'}

# Since we are setting the seed the levels will be in the same order
levels(exo_planet_data_cluster$type) <- c('hot_jupiters', 'rocky', 'others', 'cold_gas_giants')

# Cluster plot with labels
ggplot(data=exo_planet_data_cluster, aes(x=log(period), y=log(earth_radius), color=type)) + 
  geom_point() +                                                        # Adding points
  stat_ellipse(geom = "polygon", alpha = 1/2, aes(fill = type)) +       # Drawing ellipse around clusters
  labs(title="Clustering on Peroid and Earth radius",
              x = "Period", y="Earth radius") +                         # Adding labels
  theme_bw()                                                            # Adding theme
```



#### Question 8



```{r, warning=FALSE, message=FALSE, fig.align='center'}

# Binding the two dataset (data used for cluster and data with type unknown) to single one to avoid data loss
exoplanet_cmplt_data <- rbind(exo_planet_data_cluster, exo_planet_data_with_na)

# Histogram plot
ggplot(exoplanet_cmplt_data, aes(x=log(mass), fill=type)) +
  geom_histogram(colour='black') +                         # Histogram
  labs(title="Histogram of mass by type", x = "Mass") +             # Adding labels
  theme_bw()                                                        # Adding theme

# Voilin plot
ggplot(exoplanet_cmplt_data, aes(x=type, y=log(mass), fill=type)) + 
  geom_violin(trim=FALSE,  color="black")+
  labs(title="Voilin plot of mass by type",x="Type", y = "Mass")+   # Adding labels
  theme_bw()                                                        # Adding theme
```



#### Question 9


```{r, warning=FALSE, message=FALSE, fig.align='center'}

# Functing for finding the total seconds value given the a string
find_second_val <- function(time){
  as.numeric(time[1]) * 3600 + as.numeric(time[2]) * 60 + as.numeric(time[3])
}

# Splitting the r_asc and decl column with space along each row and applying converting function
r_asc = lapply(strsplit(exoplanet_cmplt_data$r_asc, " "), find_second_val)
decl = lapply(strsplit(exoplanet_cmplt_data$decl, " "), find_second_val)

# Creating tibble with the two columns
celestial_data = tibble(r_asc=unlist(r_asc), decl=unlist(decl))

# Scatter plot between the tansformed data
ggplot(
  celestial_data,
  aes(r_asc, decl)) +                                                    
  geom_point() +                                                         # Adding points
  labs(title="Celestial Map",x="Right ascension", y = "Declination") +   # Adding labels
  theme_bw()                                                             # Adding theme
```


#### Question 10


```{r, warning=FALSE, message=FALSE}
# Creating time series data
# Getting group by frequency
time_series_data = exoplanet_cmplt_data %>% group_by(meth, year) %>% summarise(Freq=n()) %>% na.omit()

# Finding cumulative frequency up to that year
time_series_data %<>% group_by(meth) %>% mutate(cum_freq = cumsum(Freq))

# Time series plot for exoplanet discovered till year
ggplot(
  time_series_data,
  aes(year, cum_freq, color=meth)) +
  geom_line() +
  labs(title = 'Times series plot for exoplanet discoverd', 
       x = 'Year', y = 'Frequency of exoplanet discovered till year') +               # Adding labels
  facet_wrap (~ meth, scales = 'free') +
  transition_reveal(year) +                                                           # Time series line
  theme_bw() +                                                                        # Adding theme
  theme(legend.position = 'None')                                                     # Removing labels
```


#### Question 11
```{r, warning=FALSE, message=FALSE}
# Define UI for application that draws a histogram
ui <- fluidPage(
   
   # Application title
   titlePanel("Exoplanet Scatterplot"),
   
   
   # Sidebar with a slider input for year and drop down menu for selecting exoplanet type
   sidebarLayout(
      sidebarPanel(
         helpText("Filter data by selecting year and type of exoplanet"),
         sliderInput("year",
                     "Select Year:",
                     min = 2009,
                     max = 2018,
                     value = 2018,
                     sep=""),
         selectInput("type", "Select Type:", 
                     choices = list("All" = "all", "Hot Jupiters" = "hot_jupiters",
                                    "Rocky" = "rocky", "Cold Gas Giants" = "cold_gas_giants", 
                                    "Unknown" = "unknown"), selected = "all")
      ),
      
      # Show a plot of the generated distribution
      mainPanel(
         plotOutput("distPlot")
      )
   )
)

# Define server logic required to draw a scatter plot
server <- function(input, output) {
   
   output$distPlot <- renderPlot({
      
      # Filtering data according to the year
      filter_data <- exoplanet_cmplt_data %>% filter(year == input$year)
      
      # Checking if the type is all if yes plotting all points
      if (input$type == "all") {
        ggplot(filter_data, aes(x=log(mass), y=log(dist), color=type)) + 
          geom_point() +
          labs(title = "Mass Vs Distance Scatter Plot", x="Mass", y="Distance") +
          theme_bw()
      }
      
      # If type is selected filtering data according to the corresponding type
      else {
        plt_title <- paste("Mass Vs Distance scatter plot for", input$type)
        filter_data <- exoplanet_cmplt_data %>% filter(year == input$year, type == input$type)
        ggplot(filter_data, aes(x=log(mass), y=log(dist))) + 
          geom_point() +
          labs(title = plt_title, x="Mass", y="Distance") +
          theme_bw()
      }
   })
}

# Run the application 
shinyApp(ui = ui, server = server, options = list(height = 500))
```



#### Question 12


```{r, warning=FALSE, message=FALSE}

# Setting options and cores
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())

# Creating model data
model_data <- exoplanet_cmplt_data  %>%  select(period, host_mass, host_temp, axis) %>% na.omit()
model_data <- log(model_data)

# Creating matrix for predictors
x <-model_data %>% select(host_mass, host_temp, axis) %>% as.matrix()

# Regression data list
exoplanet_data_lr = list(N = nrow(model_data),   # Number of rows
                     K = 3,                      # Number of predictors
                     x = x,                      # Covariates matrix
                     y = model_data$period)      # Response variable

# Maximum likelihood version
# Stan model
stan_model_lr = stan_model('lin_reg.stan')
stan_run_lr_ml = optimizing(stan_model_lr, data = exoplanet_data_lr)

# Printing summary
print(stan_run_lr_ml)
```



#### Question 13


```{r, warning=FALSE, message=FALSE}


# Sampling version
# Updated stan model
stan_model_sampling = stan_model('sampling_reg.stan')
stan_run_sampling = sampling(stan_model_sampling, data = exoplanet_data_lr)

# printing summary
print(stan_run_sampling)
```



#### Question 14

##### Parameter, Density and Trace Plot

```{r, warning=FALSE, message=FALSE}
# Plot for parameters 
plot(stan_run_sampling)

# Density plot
stan_dens(stan_run_sampling)

# Trace plot
traceplot(stan_run_sampling)
```


The first plot shows thw variance in value of parameters with the mena value and confidence. **Density Plot :** The peaks of a density Plot help display where values are concentrated over the interval. So, here the density plot shows concentrated values of the parameters alpha (Intercept) and beta. **Trace Plot :** Shows variation of parameters in each iteration.


#### Question 15

The shiny app is integrated to the R Markdown by inline using the shinyApp(). And we can give runtime as Shiny in Rmarkdown.








